Beyond PCA: UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-03-26 ms/main.tex
add disco

\templatetype{pnasresearcharticle} % Choose template
\title{UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts}

@@

Genetic structure in large cohorts results from technical, sampling and demographic variation. Visualisation is therefore a first step in most genomic analyses. However, existing data exploration methods struggle with unbalanced sampling and the many scales of population structure. We investigate an approach to dimension reduction of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to succinctly illustrate population structure in large cohorts and capture their relationships on local and global scales. Using data from large-scale genomic datasets, we demonstrate that PCA-UMAP effectively clusters closely related individuals while placing them in a global continuum of genetic variation. This approach reveals previously overlooked subpopulations within the American Hispanic population and fine-scale relationships between geography, genotypes, and phenotypes in the UK population. 
This opens new lines of investigation for demographic research and statistical genetics. Given its small computational cost, PCA-UMAP also provides a general-purpose approach to exploratory analysis in population-scale datasets.
\end{abstract}

@@

>>>>

\templatetype{pnasresearcharticle} % Choose template
\title{Beyond PCA: UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts}

@@

Genetic structure in large cohorts results from technical, sampling, and demographic variation. Identifying structure through visualization is typically the first step in most genomic analyses. Existing exploration methods struggle with unbalanced sampling and the many scales of population structure.

We investigate an approach to dimension reduction of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to succinctly illustrate population structure in large cohorts and capture their relationships on local and global scales. Using data from large-scale genomic datasets, we demonstrate that PCA-UMAP effectively clusters closely related individuals while placing them in a global continuum of genetic variation. This approach reveals previously overlooked subpopulations within the American Hispanic population and fine-scale relationships between geography, genotypes, and phenotypes in the UK population. 
This opens new lines of investigation for demographic research and statistical genetics. Given its small computational cost, PCA-UMAP also provides a general-purpose approach to exploratory analysis in population-scale datasets.
\end{abstract}

@@

%added new stuff here on cautions and how to appropriately use UMAP
Like most non-linear methods, UMAP lacks direct interpretability. The dimensions in its embeddings have no inherent meaning, unlike PCA where the axes correspond to the proportion of variance explained. UMAP is premised on local distances being more important than global distances; while points that appear very close locally can be thought to be more similar to each other, points that are very distant from each other cannot be considered to be more different from each other. Since the low dimensional embedding is formed by patching together many neighbourhoods there is a semblance of a global structure -- one that is superior to methods like t-SNE; however, the developers of UMAP caution that researchers interested primarily in global structure may wish to consider other methods.

% mention genealogical interpretation here?
% making sure I have the wording right....
% what I'd like to highlight: UMAP finds local similarities in populations. even in the HRS
% hispanic population we see a cluster appear
% basically you can find groups that seem to be uniquely closely related within a cluster
% you can't really say "how different" they are from others though
UMAP works well at highlighting groups of individuals who share more ancestry within their group, as evidenced by the Hispanic individuals in the HRS and the many 1KGP populations. When clusters form consistently, this suggests the individuals within it share more ancestry, though nothing can be said about the differences between clusters. The 1KGP dataset was deliberately crafted to come from geographically separated populations that are relatively homogeneous internally. For this reason most of its populations form isolated clusters, with the exception of those that most closely share ancestries (Iberians and Toscani, Han Chinese and Southern Han Chinese). In figure~\ref{fig:supp_megamontage_pc10_50}, population clusters bounce around in each projection, again highlighting that the positions of disconnected clusters relative to one another are not meaningful.

% with the UKBB, the geography colouring is only possible with UMAP
% tSNE would just break apart into tiny islands and scatter the people everywhere else.
UMAP excels with larger datasets containing individuals with admixed backgrounds. Unlike t-SNE, it patches together many local neighbourhoods, resulting in a relatively continuous flow to the data. t-SNE can generate seemingly arbitrary clusters, as in figures \ref{fig:supp_tsne_umap_compare_ukbb} and \ref{fig:supp_ukbb_tsne}, where small islands of ethnic groups are scattered across the projection. These are undesirable when looking at colourings by phenotypes, geography, or admixture, where we are interested in spotting gradients across populations. This continuity

We suggest making several runs of UMAP and to highlight consistent qualitative results. Tuning parameters will generally not make a difference beyond making some patterns easier to see. Selecting a lower or higher number of neighbours
